Examining Programming Students’ Interaction with Generative AI: A Metacognitive Perspective

### 2024-10-10 CSEDU2025/Paper.tex
update paper

However, researchers have also expressed concerns about the impact of these
tools on students. The systematic review by \cite{Murillo23} indicated that the
use of ChatGPT could lead to overreliance on the tool. \cite{chan23} noted that
overdependence on ChatGPT could lead to a decrease in critical thinking, as
students can make decisions based solely on the information provided by the tool.
As a potential consequence, \cite{Eason23} observed in a study that when
programming students lost access to ChatGPT-4, those who had previously relied
on it saw their performance drop by 17\%. In contrast, students who had never
used the tool were unaffected and outperformed their peers.

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54,5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding of how students interact with these tools
and how they can be used to improve learning. Furthermore, a review organized by
\cite{Lo24} emphasized the need for extended studies and objective measures to
gather more robust evidence on the use of GenAI tools in education.

@@

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{3cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise
before, during, and after cognitive activity, such as frustration, confusion,
and satisfaction, among others. Metacognitive Goals are the key to regulating
thought as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other strategies.

Spacing and interleaving are two metacognitive strategies that have been shown to
enhance learning outcomes. Spacing refers to the practice of distributing study
sessions over time, which has been shown to improve long-term retention and
understanding of the material \citep{Carvalho20}. Interleaving involves mixing
different topics or problems within a single learning session, which has been
shown to enhance long-term learning and the application of knowledge in other
situations \citep{Rivers21}.

In this study, we propose a variation of the interleaving concept, which we call
Perspective Interleaving. Unlike traditional interleaving, which typically
involves alternating between different topics or problem types, Perspective
Interleaving focuses on students' ability to shift between distinct viewpoints.
There are factors that suggest the Perspective Interleaving hypothesis can
benefit student learning. Even without an explicit alternation between different
types of problems, the student can constantly change their perspective, which
relates to the concept of contextual discrimination. Furthermore, Kolb's
Experiential Learning Theory reinforces this idea, stating that alternating
between concrete experiences and abstract reflections is essential to sustain
the learning cycle. Finally, deliberately shifting one's perspective on a
specific problem can help students reflect on their approach and improve their
metacognitive strategies.

@@

Studies on the use of metacognitive strategies indicate that these practices can
be applied to teaching the conscious use of generative AI. \cite{Zheng19}
investigated the effects of metacognitive scaffolds on group behavior,
performance, and cognitive load in computer-supported collaborative
environments. The results showed positive impacts on metacognitive
behavior and group performance without increasing cognitive load.

\cite{LiWei23} proposed a metacognition-based collaborative approach to enhance
performance in collaborative programming. The findings indicated that this
approach influenced computational thinking, critical thinking, and
metacognitive awareness.

Similarly, \cite{Wang23} explored how metacognitive instruction affects
computational thinking, critical thinking, and metacognitive skills in
collaborative programming, observing improvements in problem-solving and
collaboration.

These studies demonstrate the relevance of metacognition in enhancing
performance and skills in collaborative environments without imposing additional
cognitive load. Based on this, it is possible to suggest that metacognitive
strategies may also be useful in teaching a more conscious and critical use of
generative AI tools.

@@

\cite{Khusnul24} investigates the efficacy of a meta-learning approach in
improving metacognitive and creative skills. This study confirms the
effectiveness of meta-learning strategies and elucidates the relationship
between meta-learning and metacognition. This research suggests that
meta-learning can improve metacognitive abilities, providing valuable insights
into educational technology and course design in higher education settings.

Reconceptualizing AI Literacy: The Importance of Metacognitive Thinking in an Artificial Intelligence (AI)-Enabled Workforce

@@

CharlieBot is an educational bot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \cite{Sun24}. The RAG feature was developed using
materials from a MOOC, including text, exercises, video transcripts, and other
resources.

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, leading to the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. In this study,
students were free to use the bot as they wanted, which contributed to the
emergence of these additional categories. Table \ref{tab:categories} presents
these categories along with their respective descriptions.

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The datasets and code snippets used in this
process are available at: YYY.

@@

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. This was probably due to the fact that, in
this category, the student is more focused on understanding a concept and
engaging in deeper learning, which aligns with the metacognitive goal of
promoting comprehension and long-term retention. Figure \ref{fig:graph3}
shows the average interaction time per category.

@@

TODO: Work in progress.

* The results indicate that most interactions are focused on active learning
strategies, although deliberate use of interleaving and spacing is limited. We
believe that students can be guided \cite{Zheng19} \cite{Wang23} \cite{LiWei23}
\cite{Khusnul24} to adopt a more deliberate approach in utilizing GenAI,
enabling them to more effectively harness these tools to enhance their learning.

* Due to the anonymity of the interactions, it's unclear whether
students were intentionally using interleaving and spacing strategies or if
these patterns emerged naturally. Future research should investigate whether
students were consciously applying.

* To provide teachers with valuable data for analyzing how students are using
the tools, we are developing an AI model to automatically classify student
messages. This will enable educators to better understand student behavior and
offer more targeted support where needed.

>>>>

However, researchers have also expressed concerns about the impact of these
tools on students. The systematic review by \cite{Murillo23} indicated that the
use of ChatGPT could lead to overreliance on the tool. \cite{chan23} noted that
overdependence on ChatGPT could lead to a decrease in critical thinking, as
students can make decisions based solely on the information provided by the tool.
As a potential consequence, \cite{Bastani24} observed in a study that when
programming students lost access to ChatGPT-4, those who had previously relied
on it saw their performance drop by 17\%. In contrast, students who had never
used the tool were unaffected and outperformed their peers.

@@

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54,5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding how students interact with these tools
and how they can be used to improve learning, as emphasized \cite{Lo24}.

@@

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. \\
      Student Correction (SC) & Prompts where the student corrects the bot. \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change (LC) & Prompts that request a change of idiom. \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise before,
during, and after cognitive activity, such as frustration, confusion,
satisfaction, and others. Metacognitive Goals are the key to regulating thought
as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other possible strategies.

Spacing and Interleaving are two metacognitive strategies that have been shown
to enhance learning outcomes and supports the research questions of this study.
Spacing refers to the practice of distributing study sessions over time, which
has been shown to improve long-term retention and understanding of the material
\citep{Carvalho20}. Interleaving involve mixing different topics or problems
within a single learning session, which has been shown to also enhance long-term
learning and the application of student's knowledge in other contexts and
situations \citep{Rivers21}.

Since students are often dealing with a single problem during an interaction
session with the bot, in this study we propose a specialized form of the
Interleaving strategy, termed the \textit{Perspective Interleaving Hypothesis}.
Unlike traditional Interleaving, which typically involves alternating between
different topics or problem types, Perspective Interleaving can be characterized
by a metacognitive strategy that intentionally switches between angles of a
study object.

An important factor supporting the hypothesis is that when switching between
categories, such as an abstraction and a concrete implementation, a student is
engaging in Discriminative Contrasts \citep{kang12}. Furthermore, the hypothesis
that this approach enhances student learning may be grounded on Kolb’s
Experiential Learning Theory, which posits that learning involves a continuous
and structured cycle of concrete experiences, reflective observation, abstract
conceptualization, and active experimentation \citep{kolb14}.

@@

@@

\cite{Bastani24} investigated the impact of generative artificial intelligence
(AI), specifically GPT-4, on human learning, with a focus on mathematics
education in a high school. The problem addressed was how the use of generative
AI could affect the acquisition of new skills, which is crucial for long-term
productivity. The method involved a controlled experiment with approximately
one thousand students, who were exposed to two GPT-4-based tutors: a simple
tutor (GPT Base) and another with safeguards designed to promote learning
(GPT Tutor). The results showed that while access to GPT-4 improved performance
on practice exercises (48\% with GPT Base and 127\% with GPT Tutor), the removal
of access to GPT Base led to a 17\% decrease in student performance on exams.
This indicated that unrestricted use of *GPT Base* could hinder learning.
However, the GPT Tutor was able to mitigate this negative effect.

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes toward the technology. The study employed a
quasi-experimental design involving 82 college students divided into two
groups: one using ChatGPT-assisted programming (CFP) and the other using
traditional self-directed programming (SDP). The method combined mixed data
collection techniques, including behavioral logs, programming performance
evaluations, and interviews. The results showed that students using the CFP mode
exhibited more frequent behaviors such as debugging and reading feedback.
Although CFP students had a higher average score in programming performance,
there was no statistically significant difference compared to the SDP group
However, students’ perceptions of ChatGPT improved significantly after the
intervention, with increased perceived usefulness, ease of use, and intention to
continue using the tool in the future.

\cite{Sidra24} addressed the growing importance of metacognitive skills in an
AI-enabled workforce. The problem discussed was how AI systems, especially
collaborative tools like large language models (LLMs), impacted human
decision-making and the challenges they posed due to the lack of human-like
metacognition in AI. The method used identified four main characteristics that
differentiated human-AI interaction from human-human collaboration and proposed,
based on the dual-process theory of human thought, that metacognitive thinking
was crucial for the effective use of AI. The results indicated that human
workers needed to develop and apply metacognitive skills—such as planning,
monitoring, and evaluating their thought processes—to mitigate cognitive biases
and improve decision-making when working with AI. The study concluded by
recommending training and AI design improvements to enhance workers’
metacognitive abilities, ensuring more effective collaboration with AI systems.

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing
improvements in problem-solving and collaboration. \cite{Khusnul24} investigates
the efficacy of a meta-learning approach in improving metacognitive and creative
skills. This study confirms the effectiveness of meta-learning strategies and
elucidates the relationship between meta-learning and metacognition. This
research suggests that meta-learning can improve metacognitive abilities,
providing valuable insights into educational technology and course design in
higher education settings.

@@

CharlieBot is an educational bot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) technique to improve the precision of
responses provided to students \cite{Sun24}. The RAG approach was developed
using materials from a MOOC, including text, exercises, video transcripts, and
other resources.

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help (DH), Conceptual Question (CQ), Code
Snippet (CS), and Complete Solution (CSO). However, the data collected indicated
the need for additional categories, leading to the inclusion of four more:
Multiple Question (MQ), Student Corrections (SC), Language Change (LC), and
Uncategorized (U). In this study, students were free to use the bot as they
wanted, which contributed to the emergence of these additional categories.
Table \ref{tab:categories} presents these categories along with their respective
descriptions.

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The datasets and code used in this process are
available at: https://github.com/rodrigoprestesmachado/uc3m.

@@

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. The time gap creates an effect of
information retrieval when resuming the activity. Therefore, we can consider
this category as one in which students are more cognitively engaged. Thus, in
this category, students are likely more focused on understanding, which aligns
with the metacognitive goal of promoting comprehension and long-term retention.
Figure \ref{fig:graph3} shows the average interaction time per category.

@@

The results suggest that most student interactions with GenAI tools are centered
on active learning strategies, although the deliberate use of interleaving and
spacing techniques remains limited. Guiding students toward a more intentional
application of these strategies could enhance their learning outcomes. The
anonymity of interactions leaves it unclear whether these strategies were used
deliberately or emerged naturally, highlighting the need for future research on
students' conscious application of these methods. To support educators in
analyzing student behavior and offering more targeted assistance, an AI model
is being developed to automatically classify student messages, providing
valuable insights into how these tools are being utilized.

% TODO: Work in progress.

% * The results indicate that most interactions are focused on active learning
% strategies, although deliberate use of interleaving and spacing is limited. We
% believe that students can be guided \cite{Zheng19} \cite{Wang23} \cite{LiWei23}
% \cite{Khusnul24} to adopt a more deliberate approach in utilizing GenAI,
% enabling them to more effectively harness these tools to enhance their learning.

% * Due to the anonymity of the interactions, it's unclear whether
% students were intentionally using interleaving and spacing strategies or if
% these patterns emerged naturally. Future research should investigate whether
% students were consciously applying.

% * To provide teachers with valuable data for analyzing how students are using
% the tools, we are developing an AI model to automatically classify student
% messages. This will enable educators to better understand student behavior and
% offer more targeted support where needed.
